MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-18 docs/paper/out/paper.tex
make wording changes in using MG7, review #35 -- paper

\textbf{IDEAS}

\begin{itemize}
\tightlist
\item
  the user needs to write code, a repo would be nice, AWS account
\item
  there's some sort of conf required for the AWS account (add users,
  roles, permissions, whatever)
\item
  Why this is a good thing (or should this just be in Discussion)
\item
  something else?
\end{itemize}

@@

\subsection{Inputs, outputs, data: compile-time, expressive,
composable}\label{inputs-outputs-data-compile-time-expressive-composable}

\subsection{Tools, data, dependencies and machine
configurations}\label{tools-data-dependencies-and-machine-configurations}

\subsection{Parallel cloud execution ??}\label{parallel-cloud-execution}

@@

We use EC2, S3 and SQS through a Scala wrapper of the official
\href{https://aws.amazon.com/sdk-for-java/}{AWS Java SDK},
\href{https://github.com/ohnosequences/aws-scala-tools/releases/tag/v0.13.2}{ohnosequences/aws-scala-tools
\texttt{0.13.2}}. This uses version \texttt{1.9.25} of the AWS Java SDK.

>>>>

The MG7 16 data analysis workflow is indeed a set of tasks, all of them
based in \emph{Loquat}. For each task, a set of inputs and outputs as
well as configuration parameters must be statically defined. The user is
also free to leave the reasonable defaults for configuration, needing
only to define the input and output of the whole workflow. The
definition of this configuration is Scala code and the way of starting
an MG7 analysis is compiling the project code and launching it from the
Scala interactive console.

Code compilation prior to launching any analysis assures that no AWS
resources are launched if the analysis is not well--defined, avoiding
expenses not leading to any analysis. Besides compile--time checks,
runtime checks are made before launch to ensure existence of input data
and availability of resources.

An MG7 analysis is then a Scala project where the user only needs to set
certain variables at the code level (input, output and parameters),
compile the code and run it. To facilitate the process of setting up the
Scala project, a template with sensible defaults is provided.

In order to be able to exploit Amazon Web Services infrastructure for
the MG7 analysis, the user needs to set up an AWS account with certain
IAM (Identity and Access Management) permission policies that will grant
access to the resources used in the workflow.

@@

\subsection{Input and output data
declaration}\label{input-and-output-data-declaration}

An important aspect of the MG7 workflow is the way it deals with data
resources. All the data that is going to be used in the analysis or
produced as an output is described as Scala code using rich types from
the \emph{Datasets} language. This allows user to specify all the
information about the type of the data that can be utilized then by the
tools analyzing this data. For example, we can specify that for the
first part of the MG7 workflow running FLASH in parallel, requires
Illumina paired end reads and produces joined reads.

On one hand, specification of the input data allows us to restrict its
type and force users to be conscious about what they pass as an input.
On the other hand specification of the output data helps to build a
workflow as a \emph{composition} of several parts: we can ensure on the
Scala code type level that the output of one component fits as an input
of the next component. This can be crucial, as often the way the
analysis works depends a lot on the particular structure of the data.
For instance, in the MG7 workflow, using BLAST eDSL, we can describe
exactly which format will the output of the BLAST step have, which
information it will include, and then in the next step we can reuse this
description to easily parse BLAST output and retrieve the part of the
information needed for the taxonomy assignment analysis. Having data
structure described statically as Scala code allows us to be sure that
we won't have parsing problems or other issues with incompatible data
passed between components of the workflow.

All this doesn't compromise flexibility of the way user works with data
in MG7. On the contrary, having static data declarations as a part of
the configuration allows user to reuse component of analysis and modify
it easily according to particular needs. Besides that, an important
advantage of the type-level control is the additional insurance from
unsuccessful analysis launches, which may lead to the lost of time and
as a consequence finance spent on the cloud resources.

\subsection{Tools, data, dependencies and automated
deployment}\label{tools-data-dependencies-and-automated-deployment}

Bioinformatics software often has a complicated installation process and
requires various dependencies with unclear versions. This makes the
deployment of the bioinformatics tools an involved task and resolving it
manually is not a solution in the context of cloud computations. To face
this problem, one needs an automated system of managing tools and
resources, which will allow an expressive way for describing
dependencies between parts of a pipeline and provide a reproducible
procedure of its deployment. We have developed \emph{Statika} for this
purpose and successfully use it in MG7.

Every external tool involved in the workflow is represented as a
\emph{Statika} bundle, which is essentially a Scala project describing
the installation process of this tool and declaring dependencies on
other bundles which will be installed prior to the considered tool
itself. Describing relationships between bundles on the code level
allows us to track the directed acyclic graph of their dependencies and
linearize them to automatically install them sequentially in the right
order. Meanwhile describing installation process on the code level
allows user to utilize wide range of Scala and Java APIs, making
installation a well-defined sequence of steps rather than an unreliable
script dependent on the certain environment. This way \emph{Statika}
provides an easy way to make deployment an automated reproducible
process.

Besides the bioinformatics tools like BLAST and FLASH, \emph{Statika}
bundles are used for wrapping data dependencies and all inner components
of the system that require cloud deployment. In particular, all
components of \emph{Loquat} are bundles, which allows user to define
which components are needed for the parallel processing on each
computation unit in an expressive way, declaring them as bundle
dependencies of the loquat ``worker'' bundle. This modularization is
also important for the matter of making components of the system easily
reusable for different projects and liberating user from most of the
tasks related to their deployment.

\subsection{Parallel computations in the
cloud}\label{parallel-computations-in-the-cloud}

The MG7 workflow consists of certain steps, each of which perform some
work in parallel, using the cloud infrastructure managed by
\emph{Loquat}. It is important to notice the horizontal scalability of
this approach. Irrespectively of how much data is needed to be
proccessed, MG7 will easily handle it, by splitting data on chunks and
performing the analysis on multiple computaion units. The Amazon Elastic
Compute Cloud (EC2) provides a transparent way of managing computation
infrastructure, called autoscaling groups. User can easily set up MG7
configuration parameters, adjusting the amount of EC2 instances they
want to occupy for each task and their type. When using Amazon Web
Services for parallel computations, it is easy to scale the computation
resources, because the you pay for the time, not the amount of
computation units.

@@

MG7 uses various Amazon Web Services:

\begin{itemize}
\tightlist
\item
  EC2 (Elastic Compute Cloud) autoscaling groups for launching and
  managing computation units
\item
  S3 (Simple Storage Service) for storing input and output data
\item
  SQS (Simple Queue Service) for communication between different
  components of the system
\item
  SNS (Simple Notification Service) for notifying user about the
  progress
\end{itemize}

These services are used through a Scala wrapper of the official
\href{https://aws.amazon.com/sdk-for-java/}{AWS Java SDK
\texttt{1.9.25}},
\href{https://github.com/ohnosequences/aws-scala-tools/releases/tag/v0.13.2}{ohnosequences/aws-scala-tools
\texttt{0.13.2}}.
